---
title: "Warning message with perccalc package"
author: "Jorge Cimentada"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{"Warning message with perccalc package"}
  %\VignetteEngine{knitr::rmarkdown}
  usepackage[UTF-8]{inputenc}
---

While the other vignette shows you how to use `perccalc` appropriately, there are instances where there's just too few categories to estimate percentiles properly. Imagine estimating a distribution of `1:100` percentiles with only three ordered categories, it just sounds too far fetched.

Let's load our packages.

```{r, warning = FALSE, message = FALSE}
library(perccalc)
library(dplyr)
library(ggplot2)
```

For example, take the `survey` data on smoking habits.

```{r, message = FALSE, warning = FALSE}

smoking_data <-
  MASS::survey %>% # you will need to install the MASS package
  as_tibble() %>%
  select(Sex, Smoke, Pulse) %>%
  rename(
    gender = Sex,
    smoke = Smoke,
    pulse_rate = Pulse
  )
```

The final results is this dataset:
```{r, echo = FALSE}
smoking_data %>%
  arrange(pulse_rate)
```

Note that there's only four categories in the `smoke` variable. Let's try to estimate the percentile difference.

```{r}
smoking_data <-
  smoking_data %>%
  mutate(smoke = factor(smoke,
                        levels = c("Never", "Occas", "Regul", "Heavy"),
                        ordered = TRUE))

perc_diff(smoking_data, smoke, pulse_rate)
```

`perc_diff` returns the estimated coefficient but also warns you that it's difficult for the function to estimate the standard error. This happens similarly for `perc_dist`.

```{r}
perc_dist(smoking_data, smoke, pulse_rate) %>%
  head()
```
