---
# Example from https://joss.readthedocs.io/en/latest/submitting.html
title: 'perccalc: An R package for estimating percentiles from categorical variables'
tags:
  - R
  - categorical data analysis
  - achievement gaps
authors:
  - name: Jorge Cimentada
    orcid: 0000-0001-5594-1156
    affiliation: 1
affiliations:
 - name: Laboratory of Digital and Computational Demography, Max Planck Institute of Demographic Research (MPIDR)
   index: 1
date: 2 October 2019
bibliography: paper.bib
output: rticles::joss_article
csl: apa.csl
journal: JOSS
---

# Summary

Social science research makes extensive use of categorical variables. This
means that most variables in model definitions are a combination of categorical
and ordered categorical variables, which sometimes are proxies of continuous
variables such as income or years of education. The seriousness of this
phenomena can be best exemplified by the surge and usage of techniques tailored
specifically for this type of analysis in social science research [@agresti2010;
@agresti2011].

In particular, educational research, where there's a maturing literature on
calculating inequality gaps, categorical data are essential for estimating
inequality. For example, the income of a person is often asked in income
brackets rather than the exact amount of money; researchers would prefer the
exact amount but to avoid non-response accumulation and privacy concerns, income
brackets are a partial solution. This solution gives the income information of
respondents but at the same time in a limited fashion given that we cannot
estimate traditional statistics such as the differences of percentiles from the
income brackets. One example of this is calculating the gap in cognitive
abilities between the top (e.g 90th percentiles) and bottom (e.g 10th
percentiles) groups in the income distribution.

`perccalc` is a direct implementation of the theoretical work of @reardon2011
where it is possible to estimate the difference between two percentiles from an
ordered categorical variable. More concretely, by specifying an ordered
categorical variable and a continuous variable, this method can estimate
differences in the continuous variable between percentiles of the ordered
categorical variable. This bring forth a relevant strategy to contrast ordered
categorical variables which usually have alternative continuous measures to the
percentiles of the continuous measures. Moreover, this opens an avenue for
calculating percentile distributions and percentile differences for ordered
categorical variables which don't necessarily have an alternative continuous
measure such as job occupation classifications; one relevant example being the
classification from @erikson1979.

Recently, this method has been growing in usage in education research
[@reardon2011; @reardon_portilla2016; @bassok2016; @chmielewski2016], yet this
technique is not limited to this field alone and can be used essentially in any
context where percentiles of ordered categorical variables are of interest. One
example where this would provide useful would be in medicine based research,
where demographic characteristics such as education categories are common
factors for looking at differences between groups. Despite it's growing
popularity, there is still not an official software package that reliably
implements and tests the method in the **R** programming language [@rsoftware];
nor in any other programming language, that I'm aware of.

Currently, `perccalc` implements:

* Calculating differences in a continuous variable relative to the percentiles
  of an ordered categorical variable
* Calculating values for a continuous variable relative to the percentiles of an
  ordered categorical variable (values of a continuous variable for the 0, 10th,
  20th, ..., 100th percentile of the ordered categorical variable)
* Weight-adjusted estimations for all percentile calculations
* Provides uncertainty estimates for all calculations which allows the user to
  produce uncertainty intervals or propagate further calculations with these
  uncertainty coefficients

Below I introduce the reader to one simplified example of `perccalc`'s
capabilities in a real world scenario.

# Example

For this example, we will calculate the percentile difference in mathematics
test scores based on the education of the parents for several countries using
the PISA 2006 and PISA 2012 datasets. Let's load `perccalc` with the packages
`dplyr` [@dplyr], `tidyr` [@tidyr] and `ggplot2` [@ggplot2] for wrangling and
visualizing the data:

```{r, message = FALSE}
library(perccalc)
library(tidyr)
library(ggplot2)
library(dplyr)
```

`percalc` automatically loads `pisa_2012` and `pisa_2006` which are two datasets
with all the information that we need. These two datasets have data for Estonia,
Germany and Spain and contain the average test score in Mathematics for each
student and their father's education as measured by the international ISCED
classification. The first thing we have to do is make sure the categories are
ordered.

```{r setup}
order_edu <- c("None", "ISCED 1", "ISCED 2",
               "ISCED 3A, ISCED 4", "ISCED 3B, C",
               "ISCED 5A, 6", "ISCED 5B")

# Make ordered categories of our categorical variables
pisa_2006 <-
  pisa_2006 %>%
  mutate(father_edu = factor(father_edu, levels = order_edu, ordered = TRUE))

pisa_2012 <-
  pisa_2012 %>%
  mutate(father_edu = factor(father_edu, levels = order_edu, ordered = TRUE))

# Merge them together
pisa <- bind_rows(pisa_2006, pisa_2012)
```

Once the categories are ordered, `perc_diff` can calculate the percentile
difference between the 90th and 10th percentile for the complete sample. For
example:

```{r }
perc_diff(data_model = pisa,
          categorical_var = father_edu,
          continuous_var = avg_math,
          percentiles = c(90, 10)
          )
```

This means that the difference in Mathematics test scores between the 90th and
10th percentile of father's education is 45 points with a standard error of 14
points. We can extrapolate this example for each country separately using
`dplyr::group_by`. Since the result of `perc_diff` is a vector, to be able to
work seamlessly with `dplyr::group_by` we will use `perc_diff_df` which returns
a data frame:

```{r }
cnt_diff <-
  pisa %>%
  nest(data = c(-year, -CNT)) %>%
  mutate(
    # Look through each year/cnt combination and apply perc_diff_df
    edu_diff = lapply(data, function(x) {
      perc_diff_df(data_model = x,
                   categorical_var = father_edu,
                   continuous_var = avg_math,
                   percentiles = c(90, 10))
    })
  ) %>%
  select(-data) %>% 
  unnest(edu_diff)

title_y_axis <- "90th/10th percentiles gap in Mathematics"

# Plot results
cnt_diff %>% 
  ggplot(aes(year, difference, color = CNT, shape = CNT)) +
  geom_point() +
  geom_line() +
  theme_minimal() +
  scale_y_continuous(name = title_y_axis) +
  scale_x_continuous(name = "Year")
```

It looks like Estonia has a much smaller achievement gap relative to Spain and
Germany but also note that both Germany and Spain have been decreasing their
inequality.

`percalc` also has the `perc_dist` function which calculates the distribution of
the percentiles: this is useful for comparing more fine-grained percentiles
rather than differences:

```{r }
perc_dist(data_model = pisa,
          categorical_var = father_edu,
          continuous_var =  avg_math
          )
```

Here we get the complete father's education percentile distribution with the
test score values in mathematics for each percentile. This can be easily scaled
to all country/year combinations with our previous code:

```{r }
cnt_dist <-
  pisa %>%
  nest(data = c(-year, -CNT)) %>%
  mutate(
    edu_diff = lapply(data, function(x) perc_dist(x, father_edu, avg_math))
  ) %>%
  select(-data) %>% 
  unnest(edu_diff)

cnt_dist
```

Let's limit the distribution only to the 10th, 20th, 30th... 100th percentile
and compare for country/years:

```{r }
cnt_dist %>%
  mutate(year = as.character(year)) %>% 
  filter(percentile %in% seq(0, 100, by = 10)) %>%
  ggplot(aes(x = percentile,
             y = estimate,
             color = year,
             shape = year,
             group = percentile)) +
  geom_point() +
  geom_line(color = "black", alpha = 1/3) +
  scale_y_continuous(name = "Math test score") +
  scale_x_continuous(name = "Percentiles from father's education") +
  scale_color_discrete(name = "Year") +
  scale_shape_discrete(name = "Year") +
  facet_wrap(~ CNT) +
  theme_minimal()
```

Here the dots indicate the year 2006 and the triangles year 2012, the grey line
between them indicates the change over time. Here we can see that although
Germany and Spain are decreasing (as we saw in the plot before), the composition
of the change is very different: Spain's decrease is big all around the
distribution whereas Germany's concentrates on the top percentiles.

---

In summary, `perccalc` offers flexibility and reliability for estimating any
number of percentile differences for ordered categorical variables as well as
the distribution of percentiles values for an ordered categorical
variable. Moreover, it provides the standard errors for the estimation which can
be used to construct uncertainty intervals. This full-featured implementation
offers a reliable software to use in serious peer-review research. Researchers
can trust this implementation as an accurate representation given that it has
been built by testing it to decimal accuracy to the theoretical model of
@reardon2011; these tests are continually checked on a weekly basis making the
package particularly reliable.

The major features of `perccalc` are shown in a series of vignettes in the
package's website (https://cimentadaj.github.io/perccalc/), where there is a
direct implementation which matches @reardon2011's initial
implementation. Additionally, the package is hosted on its own open source
repository on Github (https://github.com/cimentadaj/perccalc/) and on the
official CRAN repository
(https://cran.r-project.org/web/packages/perccalc/index.html)


# Acknowledgements

I acknowledge contributions from Sean Reardon in the writing of this software
both in the form of recommendations and of the theoretical underpinnings of the
implementation.

# References
